A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)

### 2025-01-07 manuscript/manuscript.qmd
implemented SVW comments

Real-world data (RWD) is increasingly utilized to complement randomized controlled trials (RCTs) and contextualize the efficacy and safety of medical interventions through real-world evidence (RWE) studies. Simultaneously, advancements in the curation and accessibility of electronic health record data (EHR) at scale have presented the opportunity to investigate disease domains that have previously been inaccessible through administrative healthcare claims databases, including the field of oncology. The RCT DUPLICATE initiative has previously enhanced our understanding of when RWE studies can draw causal conclusions by emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas. In this process paper, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to the field of oncology. This expansion will emulate 12 RCTs in multiple EHR databases and across four different cancer indications, including non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma. The objectives of this project are to develop state-of-the-art methodological approaches and apply these to create insights that may provide guidance on the potential use of RWE in oncology. The project will place a special emphasis on a systematic evaluation of the suitability of fit-for-purpose data in relation to the study design and statistical analysis, and assessing the agreement of treatment effect estimates between RCTs and their respective emulations. Through extensive sensitivity analyses, the ENCORE project will further assess as to which measurement, design, and analytic decisions influence bias and validity in oncological RWE studies by calibrating those against RCTs.

@@

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023,[@senior2024fresh] and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs. Potential use cases comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered.

The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by comparing results against RCTs under the assumption that each RCT finding reflects a causal treatment effect. In settings where the RCT designs could be emulated well, RWE studies came to the same conclusions.[@heyard2024design] However, prior work from RCT-DUPLICATE has focused primarily on emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas using claims databases.

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project [@encoreFDA] aims to extend this work to the field of oncology which comes with its own unique set of challenges which must be systematically explored and understood. Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology is going to emulate 12 randomized clinical trials using multiple EHR data sources. The process includes an emphasis on transparency with documented assessment of data fitness of the RWD source for each trial[@rivera2024oncology; @gatto2022structured] and the conduct of extensive sensitivity analyses to assess robustness of findings and trial eligibility criteria.

The objectives of this project are to develop state-of-the-art methodological approaches and apply these to create insights that may provide guidance on the potential use of RWE for regulatory science in oncology. This includes the systematic evaluation of the suitability of data in relation to the study design and statistical analysis by emulating 12 oncology trials across four cancers and assessing the agreement of treatment effect estimates between RCTs and their respective emulations.

In this process paper, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development, study design and statistical analysis and final agreement metrics to evaluate the concordance between RCTs and emulations.

@@

The focus of ENCORE is to maximize potential learnings on when RWE studies can or cannot yield similar results compared to RCTs. To that end, the emphasis of the project is on trials of therapies for the most common cancers and/or cancers for which there has been substantial therapeutic development in recent years. After careful review and exchange with clinical and regulatory experts, four cancer indications were identified including lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer we aim to conduct three trial emulations which will be implemented using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).

The trial selection will follow a semi-automated process for which we will document the eligibility criteria resulting in a CONSORT diagram showing reasons for excluding RCTs. The search will be conducted using the AACT database which is s a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) which contains all information (protocol and result data elements) about every study registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we will use a combined search query strategy of the National Library of Medicine (NLM)-controlled *MeSH* term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.

Eligible trials need to fulfill the following basic criteria:

@@

The rationale and operationalization of each criterion is listed in detail in @tbl-criteria. We will mainly consider pivotal interventional, randomized trials after 2011 since treatment guidelines among included cancer indications have undergone significant changes in recent years. Due to the rapid adoption of new breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in the real-world. In parallel, trial results should have also not been published too recently in order to allow for enough data and follow-up time accrual in databases used for this project. A further focus is on trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol. Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] Given the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, the scope of this project was to focus on the emulation of OS as the endpoint of interest.

While most trial-eligibility criteria can be operationalized in an automated fashion, the last criterion on emulation feasibility and clinical relevance involves extensive human review. The critical points considered in this step include a thorough feasibility assessment of the data fitness [@rivera2022friends], sample size considerations and the assessment if critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and can be balanced using propensity score matching or weighting methods.[@brookhart2006variable] Lastly, trial candidates are ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4). Naturally, the majority of trials will cover advanced or metastatic cancer populations since a large proportion of drug development efforts have focused on these settings in recent years. A key learning that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD. If ongoing feasibility assessments indicate that these trials cannot be emulated with high enough confidence, runner-up candidates will be considered instead.

@@

The ENCORE project will utilize data from a total of four different oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada. All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology. A detailed description and methodology on how patients are sampled will be provided for each emulation protocol individually. For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results. If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach.

@@

For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed following the HARPER protocol template[@wang2022harmonized] and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel. Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics). All eligibility criteria will be extracted based on publicly available protocols and statistical analysis plans of the selected RCT.

@@

Real-world data fitness and emulation feasibility for a given shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care. Since a vast majority of oncological RCTs in recent years have focused on selected, biomarker-defined populations, subtleties in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population. For example, immunotherapies have significantly changed the cancer treatment landscape since the approval of the first PD-L1 inhibitor in 2015. With many trials that have followed thereafter, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has also evolved since then and PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.

@@

Critical aspects when emulating oncology trials are the choice and estimation of the appropriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the (historical) control cohort, the adoption rate of the novel intervention in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null. To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial. Example for such standard diagnostics are visualized in @fig-initiators.

The distribution of patient characteristics, stratified by exposure status, will be examined in Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT. Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, eposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting. At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design and analytic choices based on known outcome information.

@@

Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power than RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, statistical power is a critical to assess the feasibility that a comparison of agreement metrics of estimated parameters between RCT and RWD with sufficient precision can be made. Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the unstratified number of events is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic; @schoenfeld1983sample; @gsDesign]

@@

@@

For the estimation of marginal treatment effects, we will employ propensity score analytics to adjust for measured confounding between treatment arms. The selection of important prognostic covariates will be based on expert clinical knowledge and published literature on prognostic scores in oncology.[@becker2020enhanced] The implementation of propensity scores in combination with multiple imputation will follow the '*within*' methodology as described by Leyrat et al.[@leyrat2019propensity; @pishgar2020matchthem] That is, propensity score matching or weighting will be applied to each imputed dataset. The marginal treatment effect will then be estimated in each imputed and matched or weighted dataset separately and pooled into a final estimate following Rubin's rule.[@rubin2018multiple; @van2011mice] This approach has been shown to lead to unbiased estimates across different simulated scenarios with a sufficient estimation of the variance.[@leyrat2019propensity]

To asses the balance of pre-exposure covariates after matching or weighting on each imputed dataset, the average standardized mean difference (SMD) and corresponding minimum and maximum SMD range will be visualized (see example in @fig-balance). Covariate balance is typically considered at a SMD \< 0.1.[@austin2009balance] Further, we will compute the average post-matching or post-weighting C-statistics.[@franklin2014metrics] In addition, we will use a published prognostic score fro OS[@becker2020enhanced] as a balance measure to visually assess if the prognostic score is balanced between treatment arms after propensity score matching or weighting as this approach was described to show the highest correlations with bias compared with other balance measures and not affected by model misspecification.[@stuart2013prognostic]

@@

To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive two qualitative agreement metrics (regulatory agreement, estimate agreement) and one quantitative agreement based on the SMD (examples are illustrated @tbl-metrics).

**Regulatory agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance. This agreement metric mimics the question if RWE would have lead to the same regulatory conclusion as the trial, based on standard regulatory decision rules.

For the secondary endpoints of interest (e.g., median survival time or survival probabilities) only the SMD agreement metric will be applicable.

@@

Building off an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to assess the potential use of RWE for regulatory science. The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations. Historically, administrative health claims databases have been the backbone of most research in RWE. With increasing access to EHR and a maturing set of methodological approaches to draw causal inferences from such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to deliver similar causal conclusions compared to RCTs in the field of oncology.

@@

This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD. As a result, exposure misclassification (due to too narrow ascertainment windows) and selection bias (due to too long ascertainment windows) are common trade-offs in such scenarios. Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design[@gaber2024mystifying], may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters and provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.

@@

RWE based on fit-for-purpose data and principled, well-designed and reproducible studies may complement evidence coming from RCTs. The ENCORE project will systematically assess and eventually increase our understanding as to which measurement, design and analytic decisions influence bias and validity in oncological RWE studies by calibrating those against RCTs.

@@

```{r}
#| label: fig-initiators
#| fig-cap: "Visualization of descriptive drug utilization analyses displaying a) initiation trends between compared regimens based on calendar time, b) cumulative rate of patients switching to another line of treatment."
#| fig-widh: 4
#| fig-height: 7

>>>>

Real-world evidence (RWE) studies are increasingly used to complement evidence from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in clinical practice. Advancements in the curation and accessibility of electronic health record data (EHR) have presented the opportunity to investigate disease domains such as oncology, where administrative healthcare claims databases alone are not fit-for-purpose. The RCT DUPLICATE initiative has previously enhanced understanding of when RWE studies can draw causal conclusions by emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas. In this process paper, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to oncology. This expansion will emulate 12 RCTs in multiple EHR databases across four different cancer indications, including non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma. The objectives of this project are to develop state-of-the-art methodological approaches and to apply them in a demonstration project designed to create insights for guiding the use of RWE in oncology. The project will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to analysis. Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria. Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence bias and validity in oncological RWE studies.

@@

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures..

However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect. In settings where the RCT designs could be emulated well, RWE studies came to the same conclusions.[@heyard2024design] However, prior work from RCT-DUPLICATE has focused primarily on emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas using claims databases.

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project [@encoreFDA] aims to extend this work to the field of oncology. Studies in oncology come with their own unique set of challenges which must be systematically explored and understood. Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources. The process will emphasize transparency and include documented data fitness assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.

The objectives of this project are to 1) develop state-of-the-art methodological approaches and 2) to apply them to create insights that may provide guidance on the potential use of RWE for regulatory science in oncology. To achieve these objectives, this demonstration project will systematically emulate 12 oncology trials across four cancers and assess the agreement of treatment effect estimates between RCTs and their respective emulations.

In this process paper, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development and pre-registration, study design and statistical analysis, and pre-specified agreement metrics to evaluate the concordance between RCTs and emulations.

@@

The focus of ENCORE is to maximize potential learnings on when RWE studies can or cannot yield similar results compared to RCTs. To that end, the emphasis of the project is on trials of therapies for the most common cancers for which there has been substantial therapeutic development in recent years. After careful review and exchange with clinical and regulatory experts, four cancer indications were identified including lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer type we aim to conduct three trial emulations using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).

We used a semi-automated process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs for each cancer type. The search was conducted using the AACT database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains all information (protocol and result data elements) about every study registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *MeSH* term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.

Eligible trials had to fulfill the following basic criteria:

@@

The rationale and operationalization of each criterion is listed in detail in @tbl-criteria. We will mainly consider pivotal interventional, randomized trials after 2011 because treatment guidelines among included cancer indications have undergone significant changes in recent years. Due to the rapid adoption of new breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in current clinical practice. We also did not include trials with results published too recently. This decision was in order to allow for enough data and follow-up time accrual in the databases used for this project. Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] Given this, and the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest. Therefore, we include only trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.

While most trial eligibility criteria could be operationalized in an automated fashion, our final criterion was an assessment of emulation feasibility and clinical relevance. This criterion involved extensive human review. The critical points considered in this step include an initial feasibility assessment of the data fitness[@rivera2022friends], including assessment of whether critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary sample size counts are reasonable. Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4). Naturally, the majority of trials cover advanced or metastatic cancer populations because a large proportion of drug development efforts have focused on these settings in recent years. A key learning that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD. If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the sample size is not sufficient, then runner-up candidates will be considered instead.

@@

The ENCORE project will use data from a four oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada. All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology. A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results. If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects models.

@@

For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] which has been recommended for regulatory submissions of RWE studies[@guidelinegeneral], and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel. Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics). All eligibility criteria will be extracted based on publications, publicly available protocols and statistical analysis plans of the selected RCT.

@@

Real-world data fitness and emulation feasibility for each shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care. Since a vast majority of oncological RCTs in recent years have focused on selected, biomarker-defined populations, subtleties in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population. For example, immunotherapies have significantly changed the cancer treatment landscape since the approval of the first PD-L1 inhibitor in 2015. With many trials that have followed thereafter, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has also evolved since then and PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.

@@

Critical aspects when emulating oncology trials are the choice and estimation of the appropriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the (historical) control cohort, the adoption rate of the novel intervention in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null. To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial. Examples for such standard diagnostics are visualized in @fig-initiators. All exploratory analyses will be conducted blinded towards the outcome to avoid influencing study design and analytic choices.

he distribution of patient characteristics, stratified by exposure status, will be examined in Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT. Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, eposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting. At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design and analytic choices based on known outcome information.

@@

Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power than RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results. Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the unstratified number of events is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.\[@schoenfeld1981asymp\]

@@

To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: regulatory agreement, estimate agreement and agreement based on the standardized mean difference (SMD). Examples are illustrated @tbl-metrics.

**Regulatory agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.

For the secondary endpoints of interest (e.g., median survival time or survival probabilities) only the quantitative SMD agreement metric will be applicable.

@@

For the estimation of marginal treatment effects, we will employ propensity score methods to adjust for measured confounding between treatment arms. The selection of important prognostic covariates will be based on expert clinical knowledge and published literature on prognostic scores in oncology.[@becker2020enhanced] The implementation of propensity scores in combination with multiple imputation will follow the '*within*' methodology as described by Leyrat et al.[@leyrat2019propensity; @pishgar2020matchthem] That is, propensity score matching or weighting will be applied to each imputed dataset. The marginal treatment effect will then be estimated in each imputed and matched or weighted dataset separately and pooled into a final estimate following Rubin's rule.[@rubin2018multiple; @van2011mice] This approach has been shown to lead to unbiased estimates across different simulated scenarios with a sufficient estimation of the variance.[@leyrat2019propensity]

To asses the balance of pre-exposure covariates after matching or weighting on each imputed dataset, the average SMD and corresponding minimum and maximum SMD range will be visualized (see example in @fig-balance). Covariate balance is typically considered at a SMD \< 0.1.[@austin2009balance] Further, we will compute the average post-matching or post-weighting C-statistics.[@franklin2014metrics] In addition, we will use a published prognostic score for OS[@becker2020enhanced] as a balance measure to visually assess if the prognostic score is balanced between treatment arms after propensity score matching or weighting as this approach was described to show the highest correlations with bias compared with other balance measures and not affected by model misspecification.[@stuart2013prognostic]

@@

@@

Building off an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to inform the potential use of RWE for regulatory science in oncology. The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations. Historically, administrative health claims databases have been the backbone of most research in RWE. With increasing access to EHR and a maturing set of methodological approaches to draw causal inferences from such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to deliver similar causal conclusions compared to RCTs in the field of oncology.

@@

This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD. As a result, exposure misclassification (due to too narrow ascertainment windows) and selection bias (due to too long ascertainment windows) are common trade-offs in such scenarios. Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design[@gaber2024mystifying], may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters, provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.

@@

RWE based on fit-for-purpose data and principled, well-designed and reproducible studies may complement evidence coming from RCTs. Through a systematic benchmarking approach, the ENCORE project will provide insights as to how measurement, design and analytic decisions influence bias and validity in oncological RWE studies.

@@

```{r}
#| label: fig-initiators
#| fig-cap: "Visualization of descriptive drug utilization analyses displaying a) initiation trends between compared regimens based on calendar time, b) cumulative rate of patients switching to another line of treatment."
#| fig-widh: 3
#| fig-height: 6
